Replicator-mutator and noise-perturbed replicator dynamics

### 2015-01-15 paper/version_02/paper.tex
revised section 3 with smaller plots

  \includegraphics[width=\textwidth]{plots/MeanMetrics2.pdf}

@@

Interestingly, this type of variability in evolutionary outcomes can
be weeded out by impairment. To see this, we calculated the average
distance between evolved sender strategies within each group of trials
that had identical parameter values. We determined the distance
between strategies $\Sstrat$ and $\Sstrat'$ as the average Hellinger
distance between distributions $\Sstrat(\state)$ and
$\Sstrat'(\state)$ at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $\Sstrat$ with reversed message indices. The
\emph{within group distances}, i.e., the average distances between all
sender strategies in each parameter group, are plotted exemplarily in
Figure~\ref{fig:AverageEUinGroups} for $\beta = 0.1$ and $n_s =
10$. Despite minor quantitative differences the is the same for all
other parameter settings that we tested: with increasing imprecision,
the sender strategies evolving under \rdd were exactly alike (modulo
swapping of messages). This means that perceptual imprecision unifies
evolutionary outcomes and guarantees sender strategies that are not
only convex, but also regular in that they induce a vague category
split exactly in the middle of the unit interval.

@@

    \includegraphics[width=0.8\textwidth]{plots/WithinGroupMeasures.pdf}

    \caption{For each group, i.e., triplet of independent variables,
      the plots show (i) within group distances and (ii) within group
      expected utilities, as defined in the main text.}
  \label{fig:AverageEUinGroups}
\end{figure}

This unifying property of perceptual imprecision could be considered
an evolutionary beneficial side-effect. Higher imprecision can lead to
higher \emph{within group expected utility}, defined as the average
expected utility that each evolved language (i.e., pair of sender and
receiver strategy at stopping time) scored when playing against an
arbitrary other language of the same parameter
group. Figure~\ref{fig:AverageEUinGroups}) gives the representative
example of $\toler = 0.1$ and $n_s = 10$. The observation repeats for
other parameter values: while imprecision might decrease the
communicative efficiency of individual languages, it increases the
conceptual coherence and communicative success between independently
evolving strategies. It is almost as if mere confusability of states
imposes a regularity constraint on evolving categories.

>>>>

\begin{figure}[t]
  \centering

  \includegraphics[width=0.75\textwidth]{plots/MeanMetrics3.pdf}

@@

Interestingly, this type of variability in evolutionary outcomes can
be weeded out by impairment. To see this, we calculated the average
distance between evolved sender strategies within each group of trials
that had identical parameter values. We determined the distance
between strategies $\Sstrat$ and $\Sstrat'$ as the average Hellinger
distance between distributions $\Sstrat(\state)$ and
$\Sstrat'(\state)$ at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $\Sstrat$ with reversed message indices. The
\emph{within group distances}, i.e., the average distances between all
sender strategies in each parameter group, are plotted exemplarily on
the left in Figure~\ref{fig:GroupMeasures} for $\beta = 0.1$ and $n_s
= 10$. Despite minor quantitative differences the is the same for all
other parameter settings that we tested: with increasing imprecision,
the sender strategies evolving under \rdd were exactly alike (modulo
swapping of messages). This means that perceptual imprecision unifies
evolutionary outcomes and guarantees sender strategies that are not
only convex, but also regular in that they induce a vague category
split exactly in the middle of the unit interval.

@@

    \includegraphics[width=0.475\textwidth]{plots/WithinGroupDistanceConcise.pdf}
    \hfill
    \includegraphics[width=0.475\textwidth]{plots/WithinGroupEUConcise.pdf}

    \caption{Within group distance (left) and within group expected
      utility (right) as a function of impairment for all runs with
      $\toler = 0.1$ and $n_s = 10$.}
  \label{fig:GroupMeasures}
\end{figure}

This unifying property of perceptual imprecision could be considered
an evolutionary beneficial side-effect. Higher imprecision can lead to
higher \emph{within group expected utility}, defined as the average
expected utility that each evolved language (i.e., pair of sender and
receiver strategy at stopping time) scored when playing against an
arbitrary other language of the same parameter group. The right side
of Figure~\ref{fig:GroupMeasures}) gives a representative example. The
observation repeats for other parameter values: while imprecision
might decrease the communicative efficiency of individual languages,
it increases the conceptual coherence and communicative success
between independently evolving strategies. It is almost as if mere
confusability of states imposes a regularity constraint on evolving
categories.

### 2015-01-15 snippets/data_exploration/data_exploration_tmp.Rmd
revised section 3 with smaller plots

>>>>

More concise plot for second revision of paper

data$converged = data$Iterations < 201
data$convex = data$Speaker.Convex.Cat
data$entropy = (data$Speaker.entropy + data$Hearer.entropy) / 2

ds = data[,c("Number.of.states", "Impairment", "Tolerance", 
             "entropy",
             "Expected.utility",
             "convex", "converged"
             )]

dms = summarySE(dm, groupvars = c("Number.of.states", "Impairment", "Tolerance", "variable"), measurevar="value")
dsub = subset(dms, dms$Tolerance == 0.1 & dms$Number.of.states < 90)
dsub$Number.of.states = factor(dsub$Number.of.states)

myLabeller <- function(var, value){
    value <- as.character(value)
    if (var=="variable") { 
        value[value=="Expected.utility"] <- "exp. utility"
    }
    return(value)
}

pd <- position_dodge(.01)
mysize = 2.5
sp = ggplot(dsub, aes(x=Impairment, y=value, color = Number.of.states, shape=Number.of.states, linetype=Number.of.states)) + 
   geom_line(color="gray", position = pd) + xlab("imprecision") +
   geom_point(position = pd, size = mysize) +
    scale_shape_discrete(name = "# states", guide = guide_legend()) +
    scale_linetype_discrete(name = "# states", guide = guide_legend()) +
    scale_color_discrete(name = "# states", guide = guide_legend()) +
    theme(legend.position="top")

sp = sp + facet_grid(variable ~ ., scales="free", labeller=myLabeller)
show(sp)

ggsave(sp, file = "paper/version_02/plots/MeanMetrics3.pdf", width=5, height = 5.5)
